Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-05-10 papers/halfdisk/halfdisk.tex
Fixed some typos. Added reference.

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs in cartesian coordinates ($x$, $y$, and $z$) on a half-sphere  can be represented using two families of OPs on the half-disk, see \cite[Theorem 3.1]{olver2018orthogonal} for a similar construction of OPs on an arc in 2D.
%, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. 
Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

Combining each system in (\ref{eqn:jacobimatricesdefinition}) we can write the block-tridiagonal system
\begin{align*}
\renewcommand\arraystretch{1.3}
\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{pmatrix}
\bighdopab(x,y)
=
\begin{pmatrix}
	\hdopab_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{pmatrix}.
\end{align*}
where we note $\hdopab_{0,0}(x,y) \equiv \genjac_0^{(a,b)} \: P_0^{(b,b)}$, and for each $n = 0,1,2\dots$:
\begin{align*}
A_n &:= \begin{pmatrix}
		A^x_n \\
		A^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+2)} \quad (n \ne N), \quad
C_n := \begin{pmatrix}
		C^x_n \\
		C^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{pmatrix}
		B^x_n \\
		B^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{pmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align*}

@@

We can use the same technique for constructing a sparse Laplacian operator to construct the Biharmonic operator $\Delta^2$, that ensures we use a basis that satisfies both zero Dirichlet and Neumann conditions. On this note, we can also obtain the matrix operator for the Laplacian $\Delta$ that will take us from coefficients for expansion in the space $\bighdopoo(x,y)$ to coefficients in the space $\bighdop^{(2,2)}(x,y)$. We denote this matrix operator that acts on the coefficients vector as $\laplaceoo$, and is given by
\begin{align*}
    \laplaceoo := D_x^{(1,1)} \: D_x^{(0,0)} + T^{(1,2)\to(2,2)} \: D_y^{(1,1)} \: T^{(0,1)\to(1,1)} \: D_y^{(0,0)}.
\end{align*}
Further represent the Laplacian as a map from coefficients in the space $\bigW^{(2,2)}$ to coefficients in the space $\bighdopoo$. Note that a function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and Neumann boundary conditions on $\Omega$. We denote this matrix operator as $\laplacewtt$, and is given by
\begin{align*}
	\laplacewtt := W_x^{(1,1)} \: W_x^{(2,2)} + T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)} \: T_W^{(2,1)\to(1,1)} \: W_y^{(2,2)}.
\end{align*}
We can then construct a matrix operator for $\Delta^2$ that will take coefficients in the space $\bigW^{(2,2)}$ to coefficients in the space $\bighdop^{(2,2)}$. Note that any function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and zero Neumann boundary conditions on $\Omega$. The matrix operator for the Biharmonic operator is then given by
\begin{align*}
	\biharmonic = \laplaceoo \: \laplacewtt.
\end{align*}
The sparsity and structure of this biharmonic operator is seen in Figure \ref{fig:sparsity}.

@@

To obtain the recurrence coefficients for the $\{\genjac_n^{(a,b)}\}$ OPs in (\ref{eqn:Hrecurrence}), we use a version of the Stieltjes algorithm. Recall that\sotodo{check this}
\begin{align*}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x), \quad n = 0,1,2,\dots
\end{align*}
where we define $\beta_{-1}^{(a,b)} = 0$ and $H_{-1}^{(a,b)}(x) \equiv 0$. We use a variant of the Stieltjes procedure \cite{gautschi1982generating} where the polynomials are expressed as Chebyshev polynomial expansions and the inner products are calculated via Clenshaw--Curtis quadrature. This has the benefit that it is easier to incorporate high-precision arithmetic, which we use to overcome ill-conditioning present when $b$ is large, as required for large $n$ in \eqref{eq:diskpolys}. The ApproxFun.jl \cite{ApproxFun} package gives a convenient way to manipulate Chebyshev and Jacobi expansions, which we use to calculate the inner products and norms in this algorithm.

@@

We can further extend this work to trapezium shaped domains. Note that for any trapezium there exists an affine map to the canonical trapezium domain that we consider here, given by
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with 
\begin{align*}
\begin{cases}
(\alpha, \beta) &:= (0,1) \\
(\gamma, \delta) &:= (0,1) \\
\rho(x) &:= 1-\half x \\
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} = (1-x)^{a} \: x^b \: (1-\half x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b = (1-x)^a \: x^b.
\end{cases}
\end{align*}
The weight $w^{(a,b)}_P(x)$ is the weight for the shifted Jacobi polynomials on the interval $[0,1]$, and hence the corresponding OPs are the shifted Jacobi polynomials $\{\tilde{P}_n^{(a, b)}\}$. We note that the shifted Jacobi polynomials relate the the normal Jacobi polynomials by the relationship $\tilde{P}_n^{(a,b)}(x) = P_n^{(a,b)}(2x-1)$ for any degree $n = 0,1,2,\dots$ and $x \in [0,1]$. $\genjacw^{(a,b,c)}(x)$ is the (non-classical) weight for the OPs we dentote $\{\genjac_n^{(a,b,c)}\}\). Thus we arrive at the four-parameter family of 2D orthogonal polynomials $\{\hdopnk^{(a,b,c,d)}\}$ on $\Omega$ given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \tilde{P}_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) &:= \genjacw^{(a,b,c+d)}(x) \: w^{(d,c)}_P(\frac{y}{\rho(x)}) \nonumber \\
	&= (1-x)^a \: x^b \: y^c \: (1-\half x - y)^d, \quad (x,y) \in \Omega.
\end{align*}

>>>>

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs in cartesian coordinates ($x$, $y$, and $z$) on a half-sphere  can be represented using two families of OPs on the half-disk, see \cite[Theorem 3.1]{olver2018orthogonal} for a similar construction of OPs on an arc in 2D.
%, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. 
Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction \cite{staniforth2012horizontal}. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

Combining each system in (\ref{eqn:jacobimatricesdefinition}) we can write the block-tridiagonal system
\begin{align*}
\renewcommand\arraystretch{1.3}
\begin{pmatrix}
		1 & & & \\
		B_0-G_0(x,y) & A_0 & & \\
		C_1 & B_1-G_1(x,y) & \quad A_1 \quad & \\
		& C_2 & \ddots & \ddots \\
		& & \ddots &
\end{pmatrix}
\bighdopab(x,y)
=
\begin{pmatrix}
	\hdopab_{0,0} \\ 0 \\ 0 \\ \vdots \\ 0 \\
\end{pmatrix}.
\end{align*}
where we note $\hdopab_{0,0}(x,y) \equiv \genjac_0^{(a,b)} \: P_0^{(b,b)}$, and for each $n = 0,1,2\dots$:
\begin{align*}
A_n &:= \begin{pmatrix}
		A^x_n \\
		A^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+2)}, \quad
C_n := \begin{pmatrix}
		C^x_n \\
		C^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times n} \quad (n \ne 0), \nonumber \\
B_n &:= \begin{pmatrix}
		B^x_n \\
		B^y_n
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}, \quad
G_n(x,y) := \begin{pmatrix}
		xI_{n+1} \\
		yI_{n+1}
	    \end{pmatrix} \in \R^{2(n+1)\times(n+1)}.
\end{align*}

@@

We can use the same technique for constructing a sparse Laplacian operator to construct the Biharmonic operator $\Delta^2$, that ensures we use a basis that satisfies both zero Dirichlet and Neumann conditions. On this note, we can also obtain the matrix operator for the Laplacian $\Delta$ that will take us from coefficients for expansion in the space $\bighdopoo(x,y)$ to coefficients in the space $\bighdop^{(2,2)}(x,y)$. We denote this matrix operator that acts on the coefficients vector as $\laplaceoo$, and is given by
\begin{align*}
    \laplaceoo := D_x^{(1,1)} \: D_x^{(0,0)} + T^{(1,2)\to(2,2)} \: D_y^{(1,1)} \: T^{(0,1)\to(1,1)} \: D_y^{(0,0)}.
\end{align*}
Further, we can represent the Laplacian as a map from coefficients in the space $\bigW^{(2,2)}$ to coefficients in the space $\bighdopoo$. Note that a function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and Neumann boundary conditions on $\Omega$. We denote this matrix operator as $\laplacewtt$, and is given by
\begin{align*}
	\laplacewtt := W_x^{(1,1)} \: W_x^{(2,2)} + T_W^{(1,0)\to(0,0)} \: W_y^{(1,1)} \: T_W^{(2,1)\to(1,1)} \: W_y^{(2,2)}.
\end{align*}
We can then construct a matrix operator for $\Delta^2$ that will take coefficients in the space $\bigW^{(2,2)}$ to coefficients in the space $\bighdop^{(2,2)}$. Note that any function expanded in the $\bigW^{(2,2)}$ basis will satisfy both zero Dirichlet and zero Neumann boundary conditions on $\Omega$. The matrix operator for the Biharmonic operator is then given by
\begin{align*}
	\biharmonic = \laplaceoo \: \laplacewtt.
\end{align*}
The sparsity and structure of this biharmonic operator is seen in Figure \ref{fig:sparsity}.

@@

To obtain the recurrence coefficients for the $\{\genjac_n^{(a,b)}\}$ OPs in (\ref{eqn:Hrecurrence}), we use a version of the Stieltjes algorithm. Recall that\sotodo{check this}
\begin{align*}
	x \genjac_n^{(a,b)}(x) &= \beta_n^{(a,b)} \genjac_{n+1}^{(a,b)}(x) + \alpha_n^{(a,b)} \genjac_n^{(a,b)}(x) + \beta_{n-1}^{(a,b)} \genjac_{n-1}^{(a,b)}(x), \quad n = 0,1,2,\dots
\end{align*}
where we define $\beta_{-1}^{(a,b)} = 0$ and $\genjac_{-1}^{(a,b)}(x) \equiv 0$. We use a variant of the Stieltjes procedure \cite{gautschi1982generating} where the polynomials are expressed as Chebyshev polynomial expansions and the inner products are calculated via Clenshaw--Curtis quadrature. This has the benefit that it is easier to incorporate high-precision arithmetic, which we use to overcome ill-conditioning present when $b$ is large, as required for large $n$ in \eqref{eq:diskpolys}. The ApproxFun.jl \cite{ApproxFun} package gives a convenient way to manipulate Chebyshev and Jacobi expansions, which we use to calculate the inner products and norms in this algorithm.

@@

We can further extend this work to trapezium shaped domains. Note that for any trapezium there exists an affine map to the canonical trapezium domain that we consider here, given by
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with 
\begin{align*}
\begin{cases}
(\alpha, \beta) &:= (0,1) \\
(\gamma, \delta) &:= (0,1) \\
\rho(x) &:= 1-\half x \\
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} = (1-x)^{a} \: x^b \: (1-\half x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b = (1-x)^a \: x^b.
\end{cases}
\end{align*}
The weight $w^{(a,b)}_P(x)$ is the weight for the shifted Jacobi polynomials on the interval $[0,1]$, and hence the corresponding OPs are the shifted Jacobi polynomials $\{\tilde{P}_n^{(a, b)}\}$. We note that the shifted Jacobi polynomials relate to the normal Jacobi polynomials by the relationship $\tilde{P}_n^{(a,b)}(x) = P_n^{(a,b)}(2x-1)$ for any degree $n = 0,1,2,\dots$ and $x \in [0,1]$. $\genjacw^{(a,b,c)}(x)$ is the (non-classical) weight for the OPs we dentote $\{\genjac_n^{(a,b,c)}\}\). Thus we arrive at the four-parameter family of 2D orthogonal polynomials $\{\hdopnk^{(a,b,c,d)}\}$ on $\Omega$ given by, for \(0 \le k \le n, \: n = 0,1,2,\dots,\)
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: \tilde{P}_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
orthogonal with respect to the weight
\begin{align*}
	W^{(a,b,c,d)}(x,y) &:= \genjacw^{(a,b,c+d)}(x) \: w^{(d,c)}_P(\frac{y}{\rho(x)}) \nonumber \\
	&= (1-x)^a \: x^b \: y^c \: (1-\half x - y)^d, \quad (x,y) \in \Omega.
\end{align*}
